## Specified variance matrices or correlation matrices

It is sometimes useful to ...

```{r}
if (require("ade4") && require("ape")) {
    library(nlme)
    data(lizards)
    tree <- read.tree(text=lizards$hprA)
    dat <- lizards$traits[tree$tip.label, ]
    mat <- vcv(tree, corr=TRUE)# construct matrix
    fit <- gls(matur.L ~ age.mat,
               correlation=corSymm(mat[lower.tri(mat)],fixed=TRUE),
               data=dat)
```

```{r var_utils}
## FIXME: move to utils?
#' @param theta a numeric vector representing the parameters of an unstructured (general positive-definite) covariance matrix: the first n parameters are the log-standard deviations, the others are the components of the lower triangle (in rowwise matrix) of a Cholesky factor that determines the correlation matrix (see \url{http://kaskr.github.io/adcomp/classUNSTRUCTURED__CORR__t.html})
#' @return a covariance matrix
as.vcov.theta <- function(theta) {
    n <- (sqrt(8*length(theta)+1)-1)/2
    if (!isTRUE(all.equal(n,round(n))))
        stop("non-integer dimension: theta wrong length?")
    n <- round(n)
    logsd <- head(theta,n)
    cor_pars <- tail(theta,-n)
    C <- matrix(0,n,n)
    diag(C) <- 1
    ## unpack into *upper* to take advantage of R's column-major order
    C[upper.tri(C)] <- cor_pars
    Csq <- t(C) %*% C
    wt <- diag(1/sqrt(diag(Csq)))
    ret <- outer(exp(logsd),exp(logsd))*(wt %*% Csq %*% wt)
    return(ret)
}
## from Balint Tamasi, TMB user's group list
## FIXME: names based on dimnames of Sigma?
#' @param Sigma a covariance matrix
#' @param corrs.only return only values corresponding to the correlation matrix parameters?
#' @return the corresponding \code{theta} parameter vector
as.theta.vcov <- function(Sigma,corrs.only=FALSE) {
    logsd <- log(diag(Sigma))/2
    cr <- cov2cor(Sigma)
    cc <- chol(cr)
    cc <- cc %*% diag(1 / diag(cc))
    corrs <- cc[upper.tri(cc)]
    if (corrs.only) return(corrs)
    ret <- c(logsd,corrs)
    return(ret)
}
```

In this case we have to tweak a few different things:

- remove the residual variance (`dispformula=~0`; this actually sets the residual variance to a very small (non-zero) value, not zero ...)
- add the species ID to the data frame, and set up a dummy grouping variable that includes the whole data set
- fix all of the variances to be identical (in the `map` argument, assign a single factor value for the first `nsp` parameters)
- fix the correlation matrix to a known value (in the `map` argument, assign a value of `NA` for the last `length(theta)-nsp` parameters; in the `start` argument, set the corresponding elements to their fixed values

```{r}
## spp _in order_ (same as tip order)
dat$spp <- factor(rownames(dat), levels=rownames(dat))
dat$dummy <- factor(0)
nsp <- nrow(dat)
## convert correlation parameters to their corresponding parameters
corrs <- as.theta.vcov(mat, corrs.only=TRUE)
fit2 <- glmmTMB(matur.L ~ age.mat + (0+spp|dummy),
        data=dat,
        dispformula=~0,
        map=list(theta=factor(c(rep(1,nsp),rep(NA,length(corrs))))),
        start=list(theta=c(rep(0,nsp),corrs)))
## check equal coefficients to gls() fit
all.equal(c(fixef(fit2)$cond),c(coef(fit)),
          tolerance=5e-6)
## check that correlation matrix is as intended
cc <- cov2cor(VarCorr(fit2)$cond[[1]])
cc <- matrix(c(cc),
             dimnames=lapply(dimnames(cc),
                             function(x) gsub("^spp","",x)),
             nrow=nrow(cc))
all.equal(cc,mat)
```

