---
title: "Appendix C: Compare zero-inflated mixed models across R packages"
author: "Mollie Brooks and Ben Bolker"
date: "`r Sys.Date()`"
bibliography: zi.bib
bst: model5-names.bst
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
```
In this appendix, we analyze counts of begging behavior by owl nestlings. This example previously appeared in @zuur_mixed_2009 and @bolker_strategies_2013 and was originally published by @roulin_nestling_2007. The response variable is the number of calls from chicks (`NCalls`) in a nest. Since this should directly scale with the number of chicks (i.e. brood size), `logBroodSize` is used as an offset term. Since nests were repeatedly measured, `Nest` is included as a random effect. Covariates of interest include the sex of the parent visiting the nest (`SexParent`), whether the chicks were satiatited or not (`FoodTreatment`), and the timing of the parent's arrival (`ArrivalTime`).

#Preliminaries

##Load packages

```{r libs,warning=FALSE, message=FALSE,cache=FALSE}
library(glmmTMB)
library(glmmADMB)
library(MCMCglmm)
library(brms)
library(INLA)
library(broom) #for tidy
library(plyr)
library(dplyr) #tidyverse
library(ggplot2); theme_set(theme_bw())
library(ggstance)#for position_dodgev
```

## Data organization and helper functions (hidden)

```{r data}
data(Owls)
Owls = plyr::rename(Owls, c(SiblingNegotiation="NCalls"))
Owls = transform(Owls, ArrivalTime=scale(ArrivalTime, center=TRUE, scale=FALSE))
```

```{r helper,echo=FALSE}
# time 
tfun = function(...) unname(system.time(capture.output(...))["elapsed"])
abbfun = function(x) {
  gsub("[()]","", # (Intercept) -> Intercept
    gsub("(Sol\\.)*(trait|at.level\\(trait, 1\\):)*","", # simplify MCMCglmm labs
     gsub("FoodTreatment","FT",x)))  # shorten
}
tidy.brmsfit = function(x, effects=c("fixed"),...) {
    ss <- summary(x)
    tidy(ss$fixed) %>%
        mutate(effect="fixed",group="fixed") %>%
        select(effect,.rownames,Estimate,Est.Error,l.95..CI,u.95..CI,group) %>%
        rename(term=.rownames,estimate=Estimate,std.error=Est.Error,
               conf.low=l.95..CI,conf.high=u.95..CI)
}
# broom tidy method for class "inla"

tidy.inla <- function(x, effects=c("fixed"),...) {
    ss <- summary(x)
    tidy(ss$fixed) %>%
        mutate(effect="fixed",group="fixed") %>%
        select(effect,.rownames,mean,sd,X0.025quant,X0.975quant,group) %>%
        rename(term=.rownames,estimate=mean,std.error=sd,
               conf.low=X0.025quant,conf.high=X0.975quant)
}
```

# Constant zero-inflation

Here we fit the model with zero-inflation assumed to be constant across the data set, i.e. zero-inflation is independent of the predictor variables.

## glmmTMB

```{r tmbfit}
form = NCalls~(FoodTreatment + ArrivalTime) * SexParent + 
	offset(logBroodSize) + (1|Nest)
time.tmb = tfun(m1.tmb <<- glmmTMB(form,
                                  ziformula=~1, data = Owls, family=poisson))
```

## glmmADMB
```{r admbfit}
time.admb = tfun(m1.admb <<- glmmadmb(form,
                 zeroInflation=TRUE, data = Owls, family="poisson"))
```

##MCMCglmm

Code for this example was copied from @bolker_strategies_2013; a more complete description appears in the [supplementary material](https://groups.nceas.ucsb.edu/non-linear-modeling/projects/owls/WRITEUP/owls.pdf) for that paper.

```{r mcmcglmmfit}
offvec = c(1,1,2,rep(1,5)) # 1=non-offset; 2=offset
fixef2 = NCalls~trait-1+ # intercept terms for both count and binary terms
    # other fixed-effect terms only apply to count term
    at.level(trait,1):logBroodSize+
   at.level(trait,1):((FoodTreatment+ArrivalTime)*SexParent)
# residual variances independent for count and binary terms;
#    fixed to 1 for binary term
# random-effects variances independent for count and binary terms;
#    fixed very small (1e-6) for binary term
prior_overdisp  = list(R=list(V=diag(c(1,1)),nu=0.002,fix=2),   
                       G=list(list(V=diag(c(1,1e-6)),nu=0.002,fix=2)))
prior_overdisp_broodoff = c(prior_overdisp,
                              list(B=list(mu=c(0,1)[offvec],
                                V=diag(c(1e8,1e-6)[offvec]))))
time.mcmc=tfun(m1.mcmc <<- MCMCglmm(fixef2,
                                   rcov=~idh(trait):units,
                                   random=~idh(trait):Nest,
                                   prior=prior_overdisp_broodoff,
                                   data=Owls,
                                   family="zipoisson",
                                   verbose=FALSE))
```

##brms

```{r brmsfit}
time.brms = tfun(m1.brms <<- brm(form, data = Owls,
                                family="zero_inflated_poisson",
                                save_dso=TRUE))
```

```{r brmsfit2}
time.brms2 = tfun(m1.brms2 <<- update(m1.brms))
```

##INLA

```{r inlafit}
time.inla = tfun(m1.inla <<- inla(NCalls~(FoodTreatment + ArrivalTime) * SexParent + 
                                    f(Nest, model="iid"),
                              offset = logBroodSize,
                              family= "zeroinflatedpoisson1",
                              data=Owls))
```

#Comparing the results

## Timings

```{r times}
sort(c(glmmTMB=time.tmb,glmmADMB=time.admb,MCMCglmm=time.mcmc,brms=time.brms,
  brms2=time.brms2,INLA=time.inla))
```
(Time is recorded in seconds.)

`glmmTMB` fit the model in less than 5 seconds. Other methods were slower, but `MCMCglmm` was in the same order of magnitude (`brms` and `brms2` are times including and excluding compilation time, respectively).


## Estimated fixed-effect coefficients

```{r arrange_coefs,echo=FALSE, cache=FALSE}
modList = list(glmmTMB=m1.tmb,
                glmmADMB=m1.admb,
                MCMCglmm=m1.mcmc,
                brms=m1.brms,
                INLA=m1.inla)
coefs = ldply(modList,tidy,effect="fixed",conf.int=TRUE,
                  .id="model") %>%
    mutate(term=abbfun(term)) %>%
    select(model,term,estimate,conf.low,conf.high) %>%
    filter(!term %in% c("Intercept","NCalls","zi_NCalls","logBroodSize"))
```
```{r coefplot, echo=FALSE, fig.width=7, cache=FALSE}
(gg0 <- ggplot(coefs, aes(color=model,shape=model,y=term, x=estimate))+
    labs(x="Regression estimate",y="")+
    geom_pointrangeh(aes(xmax=conf.high, xmin=conf.low),
                     position = position_dodgev(height = 0.6))+
    scale_colour_brewer(palette="Set1")+
    scale_shape_manual(values=15:19))
```

*Figure C.1 -- Estimated fixed-effect coefficients* Estimates are from the same  zero-inflated Poisson model fit using functions `glmmTMB`, `glmmadmb`, `MCMCglmm`, `brm`, and `inla`.

Because we ran `brms` with flat priors, the estimates are very close to the maximum likelihood estimates of `glmmTMB`. Maximum likelihood estimates from `glmmTMB` and `glmmADMB` differ slightly because `glmmADMB` uses some numerical tricks to increase robustness and these change the objective function by a small amount.

# Complex zero-inflation

Here we fit the model with zero-inflation depending on some of the predictor variables. We can no longer use `glmmADMB` and `INLA`.

## glmmTMB

```{r tmbfit_czi}
ziform = ~FoodTreatment+(1|Nest)
time.tmb_czi = tfun(m1.tmb_czi <<- glmmTMB(form,
                            ziformula=ziform, data = Owls, family=poisson))
```

## MCMCglmm

```{r MCMCglmm_czi}
offvec_czi = c(1,1,2,rep(1,6)) # 1=non-offset; 2=offset
fixef3 = NCalls~trait-1+ # intercept terms for both count and binary terms
    # fixed-effect terms for count term
    at.level(trait,1):logBroodSize+
    at.level(trait,1):((FoodTreatment+ArrivalTime)*SexParent)+
    # fixed-effect terms for binary term
    at.level(trait,2):FoodTreatment
# residual variances independent for count and binary terms;
#    fixed to 1 for binary term
# random-effects variances now allow estimated variance for binary term
#   as well
prior_overdisp_czi  = list(R=list(V=diag(c(1,1)),nu=0.002,fix=2),   
                       G=list(list(V=diag(c(1,1)),nu=0.002)))
prior_overdisp_broodoff_czi = c(prior_overdisp_czi,
                              list(B=list(mu=c(0,1)[offvec_czi],
                                V=diag(c(1e8,1e-6)[offvec_czi]))))
time.mcmc_czi=tfun(m1.mcmc_czi <<- MCMCglmm(fixef3,
                                   rcov=~idh(trait):units,
                                   random=~idh(trait):Nest,
                                   prior=prior_overdisp_broodoff_czi,
                                   data=Owls,
                                   family="zipoisson",
                                   verbose=FALSE))
```

<!-- **to do**: don't know why we're getting the warning, we seem to be getting all of the fixed effects parameters we asked for? -->

## brms

```{r brmszifit}
time.brms_czi = tfun(m1.brms_czi <<- brm(brmsformula(form,zi=ziform),
                                 data = Owls,
                                 family="zero_inflated_poisson",
                                 save_dso=TRUE))
```

```{r brmszifit2}
time.brms_czi2 = tfun(m1.brms_czi2 <<- update(m1.brms_czi))
```

## Comparison

Timings:

```{r times_czi}
sort(c(TMB=time.tmb_czi,MCMCglmm=time.mcmc_czi,brms=time.brms_czi,
       brms2=time.brms2))
```

Coefficients:

```{r arrange_coefs_czi,echo=FALSE}
modList_czi = list(glmmTMB=m1.tmb_czi,
                MCMCglmm=m1.mcmc_czi,
                brms=m1.brms_czi)
coefs_czi = ldply(modList_czi,tidy,effect="fixed",conf.int=TRUE,
                  .id="model") %>%
    mutate(term=abbfun(term)) %>%
    select(model,term,estimate,conf.low,conf.high) %>%
    filter(!term %in% c("Intercept","NCalls","zi_NCalls","logBroodSize",
                        "zi_Intercept","zi_FTSatiated",
                        "FTDeprived:at.level, 2"))
```

```{r coefplot_czi, fig.width=7, cache=FALSE, echo=FALSE}
gg0 %+% coefs_czi
```
*Figure C.2 -- Estimated fixed-effect coefficients* Estimates are from the same  zero-inflated Poisson model with predictors on zero-inflation fit using functions `glmmTMB`, `MCMCglmm`, and `brm`.

# References
