---
title: "Priors in glmmTMB"
author: "Ben Bolker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Priors in glmmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Motivation

While `glmmTMB` is primarily designed for maximum likelihood estimation (or restricted ML), there are certain situations where it is convenient to be able to add priors for particular parameters or sets of parameters, e.g.:

* to mitigate *complete separation* (technically, when there is some linear combination of parameters that divides zero from non-zero responses in a count or binomial model; in practice, typically when one treatment combination has all-zero responses)
* to mitigate *singular fits* in random effects, i.e. when there is insufficient data to estimate a variance parameter so that single variances collapse to zero or covariance matrices are estimated with less than full rank [@chungNondegenerate2023]
* to apply a ridge penalty to a set of parameters (corresponding to an independent Gaussian prior on each parameter)

When priors are specified, `glmmTMB` will fit a *maximum a posteriori* (MAP) estimate. In other words, unlike most Bayesian estimate procedures that use Markov chain Monte Carlo to sample the entire parameter space and compute (typically)  posterior mean or median value of the parameters, `glmmTMB` will find the *mode* of the posterior distribution or the *most likely* value. The MAP estimate is theoretically less useful than the posterior mean or median, but is often a useful approximation.

It is also possible to apply `tmbstan` to a fitted `glmmTMB` model that specifies priors (see the [MCMC vignette](./mcmc.html) in order to get samples from the posterior distribution as in a more typical Bayesian analysis.

One can also think of adding priors as a way of *regularizing* or *penalizing* a maximum likelihood estimate.

## Load packages

```{r pkgs, message = FALSE}
library(glmmTMB)
library(lme4)
library(blme)
library(broom.mixed)
library(purrr)
library(dplyr)
library(ggplot2); theme_set(theme_bw())
```

### Culcita example: near-complete separation

From @bolker_glmm_2014 (regularize nearly complete separation: see more complete description [here](https://bbolker.github.io/mixedmodels-misc/ecostats_chap.html).

For comparison, we'll fit (1) unpenalized/prior-free `glmer` and `glmmTMB` models; (2) `blme::bglmer()`, which adds a prior to a `glmer` model; (3) `glmmTMB` with priors.

```{r culcita}
cdat <- readRDS(system.file("vignette_data", "culcita.rds", package = "glmmTMB"))
## identified as extreme residual in https://bbolker.github.io/mixedmodels-misc/ecostats_chap.html
cdatx <- cdat[-20,]
form <- predation~ttt+(1|block)
cmod_glmer <- glmer(form, data=cdatx, family=binomial)
cmod_bglmer <- bglmer(form, data=cdatx, family=binomial,
                    fixef.prior = normal(cov = diag(9,4)))
cmod_glmmTMB <- glmmTMB(form, data=cdatx, family=binomial)
cprior <- data.frame(prior = "normal(0, 3)",
                     class = "beta",
                     coef = "")
cmod_glmmTMB_p <- glmmTMB(form, data=cdatx, family=binomial, priors = cprior)
all.equal(coef(summary(cmod_bglmer)),
          coef(summary(cmod_glmmTMB_p))$cond,
          tolerance = 5e-2)
cmods <- ls(pattern = "cmod_[bg].*")
cmod_list <- mget(cmods) |> setNames(gsub("cmod_", "", cmods))
```

```{r culcita_comp}
cres <- (purrr::map_dfr(cmod_list,
                        ~tidy(., conf.int = TRUE, effects = "fixed"),
                        .id = "model")
    |> select(model, term, estimate, lwr = conf.low, upr = conf.high)
)
ggplot(cres, aes(x = estimate, y = term, colour = model)) +
    geom_pointrange(aes(xmin = lwr, xmax = upr),
                    position = position_dodge(width = 0.5))
```

### Gopher tortoise example: mitigate singular fit

Also from @bolker_glmm_2014:

```{r gophertortoise}
gdat <- readRDS(system.file("vignette_data", "gophertortoise.rds", package = "glmmTMB"))
form <- shells~prev+offset(log(Area))+factor(year)+(1|Site)
gmod_glmer <- glmer(form , family=poisson, data=gdat)
gmod_bglmer <- bglmer(form, family=poisson, data=gdat)
## cov.prior = gamma(shape = 2.5, rate = 0, common.scale = TRUE, posterior.scale = "sd"))
gmod_glmmTMB <- glmmTMB(form, family = poisson, data = gdat) ## 1e-5
## bglmer default corresponds to gamma(Inf, 2.5)
gprior <- data.frame(prior = "gamma(1e8, 2.5)",
                     class = "theta",
                     coef = "")
gmod_glmmTMB_p <- glmmTMB(form, family = poisson, data = gdat, priors = gprior)
vc1 <- c(VarCorr(gmod_glmmTMB_p)$cond$Site)
vc2 <- c(VarCorr(gmod_bglmer)$Site)
all.equal(vc1, vc2, tolerance = 5e-4)
gmods <- ls(pattern = "gmod_[bg].*")
gmod_list <- mget(cmods) |> setNames(gsub("gmod_", "", gmods))

```

```{r gopher_comp, eval = FALSE}
pp <- profile(gmod_bglmer, which = "theta_", devtol = Inf)
lattice::xyplot(pp)
confint(pp)
pp <- profile(gmod_glmer, which = "theta_", devtol = Inf)
cres <- (purrr::map_dfr(gmod_list,
                        ~tidy(., conf.int = TRUE, effects = "ran_pars",
                              conf.method = "profile", devtol = Inf),
                        .id = "model")
    |> select(model, term, estimate, lwr = conf.low, upr = conf.high)
)
ggplot(cres, aes(x = estimate, y = term, colour = model)) +
    geom_pointrange(aes(xmin = lwr, xmax = upr),
                    position = position_dodge(width = 0.5))
```

`blme` defaults: Wishart(dim + 2.5), or gamma(2.5). For dim = 1 (scalar), Wishart(n) corresponds to chi-squared(n), or gamma(shape = n/2, scale = n/2).  Chung et al propose `gamma(2, Inf)`; not sure why `blme` uses `gamma(2.5)` instead? or if specified via Wishart, shape = 3.5 → gamma shape of 1.75?

## To do/Fix me

- try to get internal structure of priors fixed before release, otherwise `up2date` might get annoying ...
- document synonyms
- why is `bglmer` profile CI failing (in `broom.mixed`, but not externally?)
- figure out/document `blme` default priors
- move this document to a vignette?
- add tests!
- document that gamma is applied on exp() scale
   - move prior info to a separate man page?
- implement elementwise priors
    - start with specifying by number, do lookup by name later
- allow multivariate (joint) priors on parameter vectors rather than iid priors?
     - esp for correlation matrices: LKJ, Wishart etc. (from Mikael Jagan [here](https://github.com/jaganmn/misc/tree/master/tmb_distributions))

- add info to prior structure:
   - number of prior parameters (save annoying C++ code)
   - number of model parameters to use/loop over (to be determined on R side)
       - e.g. allow 'apply gamma to all sd parameters' ?
- test!



## Development issues

It seems useful to use the API/user interface from `brms`

* downside: `brms`has lots of downstream dependencies that `glmmTMB` doesn't
    * might be able to copy the relevant code (the full [file](https://github.com/paul-buerkner/brms/blob/master/R/priors.R) is 2210 lines (!), but this includes documentation and a lot of code we don't need ...
    ```{r deps, eval = FALSE}
	
	rd <- \(x) tools::package_dependencies("brms", recursive = TRUE)[[x]]
    ## rd <- \(x) packrat:::recursivePackageDependencies(x, ignores = "", lib.loc =    .libPaths()[1])
    ## not sure why packrat and tools get different answers, but difference
    ## doesn't matter much
    brms_dep <- rd("brms")
    glmmTMB_dep <- rd("glmmTMB")
    length(setdiff(brms_dep, glmmTMB_dep))
    ```
* at its simplest, this is just a front-end for a data frame
```{r brms_priors, eval = FALSE}
## requires brms to evaluate, wanted to avoid putting it in Suggests: ...
bprior <- c(prior_string("normal(0,10)", class = "b"),
            prior(normal(1,2), class = b, coef = treat),
            prior_(~cauchy(0,2), class = ~sd,
                   group = ~subject, coef = ~Intercept))
```

```{r fake_brms_priors, echo = FALSE}
bprior <- structure(list(prior = c("normal(0,10)", "normal(1, 2)", "cauchy(0, 2)"
), class = c("b", "b", "sd"), coef = c("", "treat", "Intercept"
), group = c("", "", "subject"), resp = c("", "", ""), dpar = c("", 
"", ""), nlpar = c("", "", ""), lb = c(NA_character_, NA_character_, 
NA_character_), ub = c(NA_character_, NA_character_, NA_character_
), source = c("user", "user", "user")), row.names = c(NA, -3L
), class = c("brmsprior", "data.frame"))
```

```{r brms_prior_str}
str(bprior)
```

We probably only need to pay attention to the columns `prior`, `class`, `coef`, `group`. For our purposes, `prior` is the name and parameters; `class` will be the name of the parameter vector; `coef` will specify an index within the vector (could be a number or name?)
	
`TMB`-side data structure:

* vector of prior codes
    * we need a new `enum`, `.valid_priors`: see `make-enum` in the Makefile
* list of parameter vectors? or `prior_p1`, `prior_p2`, `prior_p3` (do any prior families have more than two parameters? What about non-scalar parameters, e.g. Wishart priors ... ???)
* vector of parameter codes (another `enum`?) (`beta`, `theta`, `thetaf` ... `b` ?)
* each index (corresponding to `coef`) is scalar, either NA (prior over all elements) or integer (a specific element)

* new loop after loglik loop to add (negative log-)prior components: loop over prior spec

## References
