---
title: "Simulate from a fitted glmmTMB model"
author: "Mollie Brooks"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
The most recent version of TMB includes the ability to simulated model, so we have implemented that ability in glmmTMB. These simulations take all random effects into account.

```{r setup, include=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs,message=FALSE}
library(glmmTMB)
library(ggplot2); theme_set(theme_bw())
```

Fit a typical model:
```{r fit1}
data(Owls)
owls_nb1 <- glmmTMB(SiblingNegotiation ~ FoodTreatment*SexParent +
                             (1|Nest)+offset(log(BroodSize)),
                          family = list(family="nbinom1",link="log"),
                          ziformula = ~1, data=Owls)
```

Then we can simulate from the fitted model with the `simulate.glmmTMB` function. It produces a list of simulated observation vectors, each of which is the same size as the original vector of observations. The default is to only simulate one vector (`nsim=1`) but we still return a list for consitency.

```{r sim}
simo=simulate(owls_nb1, seed=1)
Simdat=Owls
Simdat$SiblingNegotiation=simo[[1]]
Simdat=transform(Simdat,  
			NegPerChick = SiblingNegotiation/BroodSize, 
			type="simulated")
Owls$type = "observed"	
Dat=rbind(Owls, Simdat)	
```

Then we can plot the simulated data against the observed data to check if they are similar. Because the simulated data is conditional on the estimated random effects, the nest-specific observations should be similar to the nest-specific simulations.
```{r plots,fig.width=7}

ggplot(Dat,  aes(NegPerChick, colour=type))+geom_density()+facet_grid(FoodTreatment~SexParent)

ggplot(Dat, aes(Nest, SiblingNegotiation, colour=type))+geom_boxplot()+
 theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```
