---
title: "implementing priors in glmmTMB"
---

```{r pkgs}
library(brms)
```

It seems useful to use the API/user interface from `brms`

* downside: `brms`has 85 downstream dependencies that `glmmTMB` doesn't
    * might be able to copy the relevant code (the full [file](https://github.com/paul-buerkner/brms/blob/master/R/priors.R) is 2210 lines (!), but this includes documentation and a lot of code we don't need ...
    ```{r deps}
    rd <- \(x) packrat:::recursivePackageDependencies(x, ignores = "", lib.loc =    .libPaths()[1])
    brms_dep <- rd("brms")
    glmmTMB_dep <- rd("glmmTMB")
    length(setdiff(brms_dep, glmmTMB_dep))
    ```
* at its simplest, this is just a front-end for a data frame
```{r priors}
bprior <- c(prior_string("normal(0,10)", class = "b"),
            prior(normal(1,2), class = b, coef = treat),
            prior_(~cauchy(0,2), class = ~sd,
                   group = ~subject, coef = ~Intercept))
str(bprior)
```
We probably only need to pay attention to the columns `prior`, `class`, `coef`, `group`. For our purposes, `prior` is the name and parameters; `class` will be the name of the parameter vector; `coef` will specify an index within the vector (could be a number or name?)
	
`TMB`-side data structure:

* vector of prior codes
    * we need a new `enum`, `.valid_priors`: see `make-enum` in the Makefile
* list of parameter vectors? or `prior_p1`, `prior_p2`, `prior_p3` (do any prior families have more than two parameters? What about non-scalar parameters, e.g. Wishart priors ... ???)
* `npar` vector (number of parameters for each prior, usually 2)
* vector of parameter codes (another `enum`?) (`beta`, `theta`, `thetaf` ... `b` ?)
* each index (corresponding to `coef`) is scalar, either NA (prior over all elements) or integer (a specific element)

* new loop after loglik loop to add (negative log-)prior components: loop over prior spec
   * 




